\documentclass[11pt]{article}

\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{arxiv}

\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{graphicx}
\usepackage{booktabs}
\usepackage{listings}
\usepackage{subcaption}
\usepackage{enumitem}
\usepackage{natbib}
\usepackage{pifont}
\newcommand{\cmark}{\textcolor{codegreen}{\ding{51}}}
\newcommand{\xmark}{\textcolor{alertred}{\ding{55}}}
\definecolor{alertred}{rgb}{0.70,0.10,0.10}

\usepackage{hyperref}
\usepackage{url}
\hypersetup{
  colorlinks=true,
  linkcolor=arxivaccent,
  citecolor=arxivaccent,
  urlcolor=arxivaccent,
  filecolor=arxivaccent,
  breaklinks=true,
}

\definecolor{codebg}{rgb}{0.95,0.95,0.97}
\definecolor{codegreen}{rgb}{0.0,0.5,0.0}
\definecolor{codegray}{rgb}{0.5,0.5,0.5}
\definecolor{codepurple}{rgb}{0.58,0,0.82}
\definecolor{codeblue}{rgb}{0.0,0.0,0.7}

\lstdefinestyle{pythonstyle}{
    backgroundcolor=\color{codebg},
    basicstyle=\ttfamily\small,
    breaklines=true,
    captionpos=b,
    commentstyle=\color{codegreen},
    keywordstyle=\color{codeblue}\bfseries,
    numberstyle=\tiny\color{codegray},
    stringstyle=\color{codepurple},
    language=Python,
    showstringspaces=false,
    numbers=left,
    numbersep=5pt,
    frame=single,
    rulecolor=\color{codegray},
    tabsize=4,
    morekeywords={dataclass, List, Optional, assert, Dict, field}
}
\lstset{style=pythonstyle}

\title{RLVP: Agentic Reinforcement Learning with Verifiable Penalties\\[2pt]
\large Penalize the Path, Reward the Outcome}

\author{%
  \begin{tabular}{@{}c@{\hspace{4em}}c@{}}
    Bojie Li & Noah Shi \\
    Pine AI & University of Washington
  \end{tabular}%
}

\date{}
\runningtitle{When Do Dense Process Rewards Help Agentic RL? A Verifiable-Potential Criterion}

\begin{document}
\maketitle

\begin{abstract}
Reinforcement learning from verifiable rewards (RLVR) trains language agents on the one signal an environment checks cheaply: whether the final task succeeded. The natural way to go denser is to reward the \emph{process}---to score partial progress at each step---and a large literature builds learned or hand-designed progress rewards to do so. We argue this reaches for the wrong signal. Agentic environments are asymmetric verifiers: they cheaply and reliably confirm that an agent did something \emph{bad} (ran a destructive command, called a tool before its precondition, broke a protocol), but not that it made genuine \emph{progress}. The dense reward practitioners want lies on the hard-to-verify side; the easy-to-verify side is a \emph{penalty}.

We make the consequence precise with a within-group-variance account of group-relative RL. A group-relative advantage is, by definition, a within-group variance, and it vanishes on both the all-failing groups that dominate early training and the all-succeeding groups that dominate late. A dense process signal helps only insofar as it supplies within-group variance the outcome lacks, and only where the policy actually reaches the intermediate values that carry it. A verifiable penalty on the path meets this by construction---bad behavior is easy to sample and to check, so its signal is always reachable and un-gameable. A dense progress potential does not: on hard tasks the policy rarely reaches partial success, so the potential's variance is absent and its advertised speed-ups do not survive re-seeding.

This yields a simple recipe---\textbf{penalize the path, reward the outcome}---in which the outcome reward drives the task while a per-action verifiable penalty teaches the outcome-neutral constraints (do no harm, follow the protocol) that decide whether an agent is deployable. Across diverse tasks and model scales the recipe attains the task with near-zero violations where outcome-only training succeeds but violates on every episode, and on a real agentic benchmark it cuts harmful actions several-fold at \emph{equal} success while taking \emph{more} productive actions, not fewer.

We give four practical rules for designing the penalty, each validated by ablation: penalize verifiable \emph{actions}, not lack of progress; keep the outcome reward as the task driver and never optimize a penalty alone (a lone penalty's cheapest maximizer is to stop acting---the \emph{inaction trap}); pair the penalty with a discharge that rewards the compliant action; and make the compliant behavior reachable and its target un-gameable. We then show the mirror image: dense progress potentials, the signal the field reaches for, are fragile exactly where the variance account predicts. The contribution is a practical, mechanistic account of \emph{when and how} to spend a verifier on process signal in agentic RL---on penalties that bound the path, not potentials that promise speed.
\end{abstract}
\begin{center}
\small
Code: \url{https://github.com/19PINE-AI/rlvp} \\[2pt]
Website: \url{https://01.me/research/rlvp}
\end{center}
\vspace{-0.6em}

\vspace{-0.4em}

\input{body}

\section*{Acknowledgements}

We thank Dingkun Hong, Dongjiang Li, and Wei Zhou of ByteDance for early discussions that motivated this research.
This paper was produced using Pine Copilot's voice-directed \emph{whisper coding} workflow~\citep{pineai2025whispercoding}, in which the authors specify, discuss, and review the work by voice while a coding agent---Claude Code with Claude Opus 4.8---carries out the planning, coding, experiments, and paper writing.
We thank BSQL Networking for hosting the NVIDIA RTX PRO 6000 GPU.

\bibliographystyle{plainnat}
\bibliography{reference}

\end{document}
