% ---------------------------------------------------------------------------
% Unifying LENS (empirical, not theory). FORWARD-framed: it sits in the criterion
% section, before the experiments, so it sets them up rather than summarising them.
% One identity; each term isolated by an experiment AHEAD; robust to the method's
% per-channel-normalisation and token-placement refinements.
% ---------------------------------------------------------------------------
\subsection{A unifying account: the three gates are one variance}
\label{sec:unifying}

We stated the criterion as three gates. They are not three independent rules but three terms
of a single quantity, and the rest of the paper is in effect a tour of it---each experiment
ahead isolating one term. We give the identity not as a theorem but as the picture to hold
while reading the evidence.

Group-relative methods learn from a group of $G$ rollouts only through how its rewards
\emph{differ}: each reward is centred on the group mean, $A_i = R_i - \bar R$, so identical
rewards make every advantage zero and the group yields no update, whatever the rollouts
did.\footnote{Standard GRPO also divides the advantages by the group standard deviation; this
rescales the non-zero case but not \emph{when} the advantages vanish.} Summarise a rollout by
its outcome plus a weighted verifiable potential, $R_i = O_i + \beta\,\Phi_i$, with
$O_i\in\{0,1\}$ and $\Phi_i$ a scalar process score (goals discharged, stages completed, tests
passed). The spread the group exposes is then a variance of a sum:
\[
  \operatorname{Var}_G(R)\;=\;\operatorname{Var}_G(O)\;+\;\beta^2\operatorname{Var}_G(\Phi)
  \;+\;2\beta\,\operatorname{Cov}_G(O,\Phi).
\]
The identity is trivial; what is not is that its three terms are the three gates, each
isolated by a different experiment ahead:

\begin{itemize}[leftmargin=1.4em,itemsep=2pt,topsep=2pt]
\item \textbf{Outcome term -- all-fail blindness.} When every rollout fails---common early on
a hard task---$O_i\equiv 0$, so this term and the cross term vanish; with no potential
($\beta=0$) the group is silent. The theorem-proving efficiency result (\S\ref{sec:efficiency})
lives in this regime.
\item \textbf{Potential term -- finer and reachable.} The process channel adds signal only
when $\operatorname{Var}_G(\Phi)>0$, which collapses two of the criterion's gates---finer and
reachable---into one quantity: $\Phi$ must \emph{be able} to take more than one value (finer
than the binary outcome) and the policy must \emph{actually reach} them. A potential equal to
the outcome, or one never moved off zero, contributes nothing---the software-repair null
(\S\ref{sec:law}), where $\Phi=0$ on all $156$ rollouts.
\item \textbf{Cross term -- un-gameable.} A non-zero potential variance still need not point
\emph{toward} success; the covariance is what says it does. A signal whose variation is
uncorrelated with the outcome optimises something else---the compliance collapse of a
misaligned penalty (\S\ref{sec:law}).
\end{itemize}

We claim no more than the identity earns. It is not a proof that a process reward helps---the
cross-term reading especially concerns training \emph{dynamics}, which only the experiments
settle. And the method's two refinements, per-channel normalisation and per-token placement
(\S\ref{sec:anatomy}), rescale and route this signal; reassuringly, they leave intact the only
structure the lens depends on---which terms vanish, and the sign of the alignment. What the
identity buys is a sentence to carry into a new domain---\textbf{a useful dense process reward
is the within-group variance the outcome cannot supply: non-zero when finer and reachable,
aligned when un-gameable}---and a quantity, $\operatorname{Var}_G(\Phi)$, cheap to measure on a
handful of base rollouts before any training. The phase diagram below does exactly that, and
finds the benefit of a dense reward rising with it.
